---
title: "Example gene-wise analysis of the 1001 genomes data set"
author: "R. Clay Wright"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This web app and open source software package for the R programming language was built to facilitate exploration of the naturally occuring sequence diversity for your favorite gene or gene family. We hope this product will help plant biologists without bioinformatics experience access the 1001 genomes project’s rich catalog of natural variation, formulate hypotheses and identify new alleles in their favorite gene family.

The 1001 genomes dataset was collected with the intention of genome scale analyses, such as genome wide association studies, in which a particular phenotype is quantified for all of the accessions and sequence variants that are associated with phenotypic variation are identified. In my recent work, I have utilized this rich dataset in a different manner to measure the effects of coding sequence variation on protein functional variation and phenotypic variation. This work has expanded the sequence-function-phenotype map for the auxin-signaling F-box gene family, and this product will facilitate similar analysis other genes, gene families, and genomic regions.



So pick out your favorite gene, gene family or genomic region and follow the steps in this guide to analyze its natural variation.

## 




Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
